---
title: "Add a Title Here"
author: "Your Name Here"
date: "2022-09-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "..")
```

## Introduction

Give a brief introduction of the research question(s), and how you will answer them.

## Data

Import here the data and describe it. Include in the chunk below a brief summary

```{r data, echo = FALSE}
# copy paste the code to import the data
data <- read.csv("Data/height.csv")
#knitr::kable() #try to copy and paste here a relevant line of code as 
# an argument to the kable() function.
```

## Analysis

Describe how we analysed the data. You might wish to add a snippet of code within your analysis, such as `r round(mean(data$height), 2)`. 

Include below a relevant visualisation for the mean. 

```{r data visualisation, echo = FALSE}

```

Let's also include some analysis of whether men and women are, on average, of different heights. We might add the output of the t test in a chunk

```{r t test}

```

And also a visualisation - the box plot would be good here.

```{r box plot, echo = FALSE}

```

## Conclusion

Add a few words in conclusion. What weaknesses might there be in your research? What further questions might you answer?